Discovery and analysis of pleiotropic eQTLs

### 2023-01-20 presentation_230120_internal_seminar/intern_sem_230120.tex
ENH

\title{Analysis of pleiotropy of eQTLs}
\subtitle{Internal Seminar}
\author{Aitor Gonz\'alez}
\institute{Aix Marseille Univ, INSERM, TAGC}
\date{Jan. 20, 2023}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    The associations to diseases are heterougeneously distributed in the genome, Watanabe & Posthuma
%    This is based on loci associated to multiple traits
    \begin{frame}
        \frametitle{Genome-wide association studies (GWAS)}

        \begin{itemize}
            \item Association test of effect allele on case individuals
            \item High-throughput genetics: 3000 unique loci for over 250 disease/trait outcomes\footnote{Loos 2020. 10.1038/s41467-020-19653-5}
            \item Of the 1,707.0 Mb genome, 90.0\% is associated to different GWAS categories
        \end{itemize}
%
        \vfill

%          However, how do we know if these variants are active in a tissue,
%         or whether these variants false positives due to linkage disequilibrium,
%         or how these variants act molecularly
        Questions:\footnote{Cano-Gamez et al. 2020. doi:10.3389/fgene.2020.00424}
%
        \begin{itemize}
            \item Correlation between variants. Which is the causal variant?
            \item Which cell types important for disease?
            \item In which tissues are these variants active?
        \end{itemize}
%
        \vfill
%
        Questions partially addressed by looking at eQTLs in different tissues

@@

        \begin{enumerate}
            \item To identify pleiotropic eQTLs in the genome
            \item To characterise their genomic regions, target genes and the cause of their pleiotropy
        \end{enumerate}
    \end{frame}

@@

        \begin{itemize}
            \item 418 GWAS
            \item Physiological and disease traits
            \item UK biobank, GWAS catalog, Other datasets
        \end{itemize}

@@

        \begin{itemize}
            \item 127 studies in human from different datasets
            \item Tissues, cell types, stimulated conditions, ...
            \item GTEX, DICE, ...
        \end{itemize}

@@

        \begin{center}
            \includegraphics[width=\textwidth]{../presentation_230120_gold2022_paris/fig/pgen.1004383.g001.png}
        \end{center}

@@

@@

        \begin{itemize}
            \item Colocalization between 293 GWAS and 127 eQTL studies
            \item 25\% of 10 k leading SNPs (SNP.PP.H4$\geq$0.5), colocalized with at least one eQTL
            \item At SNP.PP.H4$\geq$0.5, 5 k eQTLs and 5 k target genes
            \item 138e3 colocalized variants with probability $\geq$ 0.75
            \item Percentage of GWAS loci explained by eQTL: 0$\%$ - 77$\%$ (Average 33$\%$)
        \end{itemize}

@@

        \begin{itemize}
            \item Uniformly annotate GWAS traits based on clustering and EBI ontologies
            \item Aggregate colocalised eQTLs based on the number of trait categories
            \item Analyse eQTLs in groups based on the number of trait categories
        \end{itemize}

@@

        \begin{itemize}
            \item Example eQTLs involved in four GWAS categories in different cytobands
            \item The representative gene is the eQTL gene with the highest PubMed publication count
        \end{itemize}

@@

        \begin{itemize}
            \item Cell cycle regulator
            \item GWAS categories: cancer;cardiovascular disease;hypertensive disorder;joint disease;respiratory system disease;skin cancer
            \item eQTL biosamples: adipose tissue;blood;brain;colon;immune system; ...
        \end{itemize}

@@

        \begin{itemize}
            \item Cell cycle regulator
            \item GWAS categories: cancer;cardiovascular disease;hypertensive disorder;joint disease;respiratory system disease;skin cancer
            \item eQTL biosamples: adipose tissue;adrenal gland;aorta;blood;brain; ...
        \end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Where are the pleiotropic variants?\footnote{https://www.ensembl.org/vep}}

        \begin{center}
            \includegraphics[width=0.7\textwidth]{\floatRelativePath/pltbar_vep_consequence.py/vep.png}
        \end{center}

        In the transcripts

%        \let\thefootnote\relax\footnotetext{$^1$https://www.ensembl.org/vep}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Are pleiotropic variants associated to more eQTL genes?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Are pleiotropic variants associated to more eQTL tissues?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Do pleiotropic variants bind more transcription factors?\footnote{https://remap.univ-amu.fr}}

@@

>>>>

\title{Discovery and analysis of pleiotropic eQTLs}
\subtitle{Internal Seminar}
\author{Aitor Gonz\'alez}
\institute{Aix Marseille Univ, INSERM, TAGC}
\date{Jan. 20, 2023}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    The associations to diseases are heterougeneously distributed in the genome, Watanabe & Posthuma
%    This is based on loci associated to multiple traits
    \begin{frame}
        \frametitle{GWAS and pleiotropy}

        \begin{itemize}
            \item Genome-wise association studies: Association test in the population of variants on common traits, eg. height or Crohn's disease
            \item Used for frequent variants and common quantitative or qualitative traits
            \item High-throughput genetics: 3000 unique loci for over thousands of disease/trait outcomes\footnote{Loos 2020. 10.1038/s41467-020-19653-5}
            \item Of the 1,707.0 Mb genome, 90.0\% is associated to different GWAS categories\footnote{Watanabe et al 2019. 10.1038/s41588-019-0481-0}
        \end{itemize}
%
%         \vfill

%          However, how do we know if these variants are active in a tissue,
%         or whether these variants false positives due to linkage disequilibrium,
%         or how these variants act molecularly
%         Questions:\footnote{Cano-Gamez et al. 2020. doi:10.3389/fgene.2020.00424}
% %
%         \begin{itemize}
%             \item Correlation between variants. Which is the causal variant?
%             \item Which genes are important for disease?
%             \item In which tissues are these variants active?
%         \end{itemize}
% %
%         \vfill
% %
%         Questions partially addressed by looking at eQTLs in different tissues

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    The associations to diseases are heterougeneously distributed in the genome, Watanabe & Posthuma
%    This is based on loci associated to multiple traits
    \begin{frame}
        \frametitle{Questions regarding pleiotropy analysis\footnote{Cano-Gamez et al. 2020. doi:10.3389/fgene.2020.00424}}

%         \begin{itemize}
%             \item Association test of effect allele of individuals on traits
%             \item High-throughput genetics: 3000 unique loci for over 250 disease/trait outcomes\footnote{Loos 2020. 10.1038/s41467-020-19653-5}
%             \item Of the 1,707.0 Mb genome, 90.0\% is associated to different GWAS categories\footnote{Watanabe et al 2019. 10.1038/s41588-019-0481-0}
%         \end{itemize}
% %
%         \vfill

%          However, how do we know if these variants are active in a tissue,
%         or whether these variants false positives due to linkage disequilibrium,
%         or how these variants act molecularly

%
        \begin{itemize}
            \item Correlation between variants. Which is the causal variant?
            \item Which genes are important for disease?
            \item In which tissues are these variants active?
        \end{itemize}
%
        \vfill
%
        Questions partially addressed by looking at eQTLs in different tissues

@@

        \begin{enumerate}
            \item To identify pleiotropic eQTLs in the genome
            \item To characterise the causes of their pleiotropy
        \end{enumerate}
    \end{frame}

@@

        \begin{itemize}
            \item UK biobank, GWAS catalog, Other datasets
            \item Physiological and disease traits
            \item 418 GWAS, based on sample size and traits
        \end{itemize}

@@

        \begin{itemize}
            \item GTEX, DICE, ...
            \item Tissues, cell types, stimulated conditions, ...
            \item 127 studies in human from different datasets
        \end{itemize}

@@

        \begin{center}
            \includegraphics[width=0.5\textwidth]{../presentation_230120_gold2022_paris/fig/pgen.1004383.g001.png}
        \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Colocalization of GWAS and eQTLs}

        \begin{itemize}
            \item Colocalization between 293 GWAS and 127 eQTL studies
        \end{itemize}

@@

        \begin{itemize}
            \item Colocalization between 293 GWAS and 127 eQTL studies
            \item 25\%\footnote{SNP.PP.H4$\geq$0.5} of 10,000 leading SNPs, colocalized with at least one eQTL
            \item 5,000 eQTLs and 5,000 target genes\footnote{SNP.PP.H4$\geq$0.5}
            \item 138e3 colocalized variants with probability $\geq$ 0.75
            \item Percentage of GWAS loci explained by eQTL: histogram
        \end{itemize}

@@

        \begin{itemize}
            \item Annotate GWAS traits uniformly with ontologies
            \item Annotate normalized trait name in higher-level categories, eg cardiovascular
            \item Group categories of each eQTL
            \item Analyse eQTLs in groups based on the number of trait categories
        \end{itemize}

@@

        \begin{itemize}
            \item Example pleiotropic eQTLs in different cytobands
            \item The representative gene is the eQTL gene with the highest PubMed publication count
        \end{itemize}

@@

        \begin{itemize}
            \item Cell cycle regulator
            \item Trait categories: cancer;cardiovascular disease;hypertensive disorder;joint disease;respiratory system disease;skin cancer
            \item eQTL biosamples: adipose tissue;blood;brain;colon;immune system; ...
        \end{itemize}

@@

        \begin{figure}[!]
            \includegraphics[width=0.95\textwidth]{fig/tp53}
        \end{figure}

        \begin{itemize}
            \item Cell cycle regulator
            \item Trait categories: cancer;cardiovascular disease;hypertensive disorder;joint disease;respiratory system disease;skin cancer
            \item eQTL biosamples: adipose tissue;adrenal gland;aorta;blood;brain; ...
        \end{itemize}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%     \begin{frame}
%         \frametitle{Where are the pleiotropic variants?\footnote{https://www.ensembl.org/vep}}
%
%         \begin{center}
%             \includegraphics[width=0.7\textwidth]{fig/vep.png}
%         \end{center}
%
%         In the transcripts
%
% %        \let\thefootnote\relax\footnotetext{$^1$https://www.ensembl.org/vep}
%
%     \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Are pleiotropic eQTLs associated to more eQTL genes?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Are pleiotropic eQTLs associated to more eQTL tissues?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Do pleiotropic eQTLs bind more transcription factors?\footnote{https://remap.univ-amu.fr}}

@@

        \url{https://gwas2eqtl.tagc.univ-amu.fr/gwas2eqtl/}
